Reversible Regime Change: Climate-driven phytoplankton community shifts in the Cariaco Basin, Venezuela

### 2025-12-23 agujournaltemplate.tex
Start of Reviewer #2 responding

Phytoplankton communities are integral to oceanic biogeochemical cycles and are sensitive indicators of climate-driven environmental variability.
Long-term time series capture this variability, allowing us to unravel the effects of environmental change on local communities. This study investigates changes in the phytoplankton community of the Cariaco Basin, off the coast of Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data. To investigate impacts of larger-scale climate variability, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both linked to variations in local weather and upwelling in the Cariaco Basin.
Cluster analysis identified two distinct community states separated by a shift from high to low upwelling conditions in 2004. In 2014, community and environmental conditions returned a state similar to that before the shift in 2004. In contrast, diversity declined between 1998 and 2009 independently of changes in the upwelling regime.
To quantify effects of climate and bottom-up drivers on phytoplankton community changes, we employed gradient forest analysis and found the AMO index was the strongest predictor, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2. Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and suggests a possible cyclic (if not resilient) behavior of this tropical marine ecosystem under global change.

\end{abstract}

@@

% PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
The shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensities from 2004 to 2013 did not appear to be a permanent change in regime. Data from 2014 to 2017, the final years of the CARIACO (CArbon Retention In A Colored Ocean) time series, indicated a return to strong upwelling intensities and high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}. The impact of large-scale climatic processes on the observed shifts in the phytoplankton community remains unclear. The Atlantic Multidecadal Oscillation (AMO) has emerged as an important factor influencing the climate system in the region and the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean Sea \cite{stephenson_changes_2014}. The El Niño Southern Oscillation (ENSO) is teleconnected to the Caribbean Sea, given that during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures increase and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. However, using the time series available only until 2010, \citeA{taylor_ecosystem_2012} found no significant correlation between MEI v.2 and in situ observations, and did not test for correlations with the AMO index.

% What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
\citeA{taylor_ecosystem_2012} noted that the observed changes could not be clearly identified as ``unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation''. To our knowledge, all analyses of phytoplankton community diversity in the Cariaco Basin do not include time series data beyond 2012, and no detailed investigations have yet been conducted on changes in diversity based on microscopy cell counts. Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 to 2017, 2) changes in the structure of the phytoplankton community and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community. Our analysis focuses exclusively on bottom-up drivers of community changes because these data are available throughout the entire time series and at a monthly resolution. Our study builds on and expands the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. In doing so, we aim to provide a more comprehensive assessment of long-term phytoplankton community dynamics in the Cariaco Basin, offering new insights into the local ecological responses to environmental variability across decadal scales.

@@

\subsection{Meteorological and climate data}
    %from Taylor et al. 2012
    In addition to the in situ data, we also examined meteorological and climatic data that are known to influence ocean conditions and marine ecosystems. Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}. We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the monthly averaged ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the CARIACO time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size). For simplicity, we refer to this variable as ``wind speed'' throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate of total precipitation, comprising all water that falls on Earth's surface as rain or snow.

    In addition to the wind data, we examined two climatic indices, each of which provides information on the basin-scale climatic conditions and, consequently, the broad-scale forcing of marine ecosystems. The monthly Atlantic Multidecadal Oscillation (AMO) index data (10-year low-pass), which represent area-averaged, detrended low-pass filtered North Atlantic SST anomalies (\ang{0}N-\ang{60}N) \cite{trenberth2023amo}.

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP; SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W) \cite{noaa_psl_meiv2}. In our analysis, the bi-monthly values were aligned with the tail month (December-January is treated as the value for January).

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using ``tidyverse'' packages \cite{wickham_welcome_2019}. All the codes we produced for data processing and data analysis are publicly available on GitHub \cite{benjamin_post_2025_16878343}.
    For all discrete depth measurements, the data were interpolated using the ``oce'' package \cite{kelley_oce_2023} with the adaptive ``unesco'' algorithm defined by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to \qty{100}{\meter} and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis. Furthermore, the sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}. For phytoplankton microscopy data, individually identified taxonomic units (species or genus) were interpolated and integrated across depth, and finally grouped by genus via the sum of interpolated counts.

\subsection{Data analysis}    
    In the analysis, we set out to find if there was a structure in the distribution and temporal evolution of diversity metrics and community composition and, additionally, to identify potential bottom-up drivers of this change. 
    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores, $z = \frac{(x – \mu)}{\sigma}$, where $x$ is the yearly mean, $\mu$ is the mean and $\sigma$ is the standard deviation of all yearly means. For the analysis of annual z-scores, data from 1995, 2016, and 2017 were not included because insufficient coverage (i.e. more than six missing monthly samples) in these years prevented us from producing reliable yearly estimates.

@@

\subsection{AMO and upwelling-related variables drive community changes}
    % Figure 6
    To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a gradient forest analysis on the monthly microscopy cell counts interpolated over the top 100 meters and grouped by identified genus. Preliminary model runs were made to determine whether time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months. Figure \ref{fig:GF} $a$ shows the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity. The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentration scored the lowest in predicting changes within the community. 
    Gradient forest predicts shifts in the community by building a random forest model for each observed genus across the time series. The predictive power of the models for each genus is shown in Figure \ref{fig:GF} $b$. The microscopy data are biased toward larger cell sizes and, therefore, most of the predicted genus cell counts are diatoms. However, two haptophyte and three dinoflagellate genera rank relatively high. For the presented run a total of 89 genera (each observed more than 10 times over the time series) were used as indicator variables, of which 46 genera could be predicted using the given predictor variables, which corresponds to more than \qty{95}{\%} of counted cells (see Table \ref{tab:sup:GenusFuncGroups} for a complete list of identified genera). 
    Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the parameter range. The black lines indicate cumulative importance across the entire community, and the colored lines indicate importance per phytoplankton group. 
    Cumulative importance curves from gradient forest analysis show how the importance of a predictor accumulates across its range of values. The slope of the curve at any point indicates where the strongest compositional turnover occurs—steep sections represent threshold values where small changes in the predictor correspond to large changes in community composition, while flat sections indicate regions where the predictor has little influence on community structure.
    For the AMO index, we see a clear pattern in which negative anomalies affect the community (slope is highest for negative values). This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes and cyanobacteria. Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range within which the community changed most, indicating these values as a threshold for community composition. Figure \ref{fig:sup:GFoutput_lags_extra} in Appendix shows response curves for all variables not shown here. Variables that show a pronounced threshold response are evaporation and salinity. For evaporation the threshold is at larger values of evaporation, which are most likely linked to periods of strong stratification, and salinity at both extreme ends of the range, signifying effects on cell counts of diatoms, dinoflagellates and cyanobacteria at strong upwelling and stratified conditions.

    \begin{figure}
    \noindent\includegraphics[width=\textwidth]{fig/Figure6_GF_output_FINAL_v2.pdf}
    \caption{Gradient forest analysis of monthly community data at a genus level. (a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances. (b) Phytoplankton genera that could be predicted ranked according to the predictive power via the weighted importance. (c) Cumulative importance curves for the predictor variables. Black line shows cumulative importance across predictor range for the entire model and all genera. Dashed colored lines represent the responses of individual functional groups.}
    \label{fig:GF}
    \end{figure}

@@

    % P3: Shift in Diversity does not match shift in biomass/cell counts, why?
    The temporal dynamics found in yearly mean genus richness, however, did not correspond directly to the observed clusters. Instead, there was a marked reduction from 1998 to 2009, after which richness recovered four years prior to 2014, when the community returned to levels characteristic of cluster 1. The same trend was evident, albeit to a lesser degree, in the other diversity metrics. Generally, mean annual anomalies in Shannon diversity and Pielou's evenness correlated with genus richness and exhibited a reduced yearly mean between 2000 and 2006, during which cell counts recorded an increase in nanoflagellates and a pronounced peak in cyanobacteria.
    We did not find a significant difference in diversity between the two clusters, either by Shannon index or by genus richness (Figures \ref{fig:zscore} and \ref{fig:divts}). However, we did observe a significant difference in Pielou's index between the two clusters. The results suggest that the phytoplankton community became more evenly distributed during the low upwelling conditions of cluster 2, which is consistent with the analysis of \citeA{pinckney_phytoplankton_2015} based on HPLC pigment data. This increase in evenness, quantified by the Pielou index, is most likely explained by the lack of strong upwelling events and the resulting reduced disturbance of the community. The resulting decrease in compositional turnover was also observed in a decline in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$). As the productivity and biomass of phytoplankton communities are driven primarily by nutrients supplied through disturbance events, this is in line with the inverse relationship between evenness and community performance observed in previous studies \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.

@@

    The El Niño Southern Oscillation (ENSO) exhibits a strong teleconnection to the Caribbean Sea. During El Niño (indicated by a positive MEI v.2 index) sea surface temperatures rise, and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. This would imply a general reduction in upwelling strength during El Niño events in the Cariaco Basin, which has indeed been observed during the period from 1997 to 1998 \cite{romero_seasonal_2009}. The MEI v.2 is significantly positively correlated to local evaporation and salinity and negatively correlated to the AMO index, however it is not significantly correlated to local upwelling-related variables (see Figure \ref{fig:sup:correlation} in Appendix). The link here is less clear; while generally trade wind strength and SST are anti-correlated in the Caribbean Sea, positive trade wind anomalies have been observed during during El Niño conditions and the reverse during La Niña \cite{chang_coupled_2013}. \citeA{taylor_ecosystem_2012} found no strong correlation between MEI v.2 and other variables in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance using linear regression with a time lag of 12 months. In contrast, we found a relatively strong predictive power for changes in the phytoplankton community for the MEI v.2 index with a 4-month time lag. We also tested a 12-month time lag and found no significant improvement (Table \ref{tab:sup:GFlagtests} in Appendix).
    The threshold responses indicate a shift in the abundances of dinoflagellates for negative MEI v.2 anomalies, which represent La Niña conditions. Findings from Bayesian modeling showed that dinoflagellates responding positively to warmer, stratified conditions in the Cariaco Basin \cite{mutshinda_environmental_2013}. For cyanobacteria and haptophytes, we observed a pronounced threshold response to positive anomalies, demonstrating sensitivity to ENSO events. Diatoms do not exhibit a clear threshold response, but show a slightly stronger response in the positive range of the MEI v.2. Interestingly, positive anomalies in the MEI v.2, representing strong ENSO events, are mostly found in the high upwelling regime of cluster 1 (see Figure \ref{fig:clustering} $e$). This could be explained by previous findings, that strong El Niño events (as indicated by a positive MEI) increase the strength of currents in the Caribbean sea through modulating the easterly wind anomalies, thereby potentially increasing upwelling strength in the Cariaco Basin \cite{huang_enso_2023}.

@@

    These results matched the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, as the scaling axis that separated the two clusters was NMDS 1, which co-varied most strongly with changes in AMO, MEI v.2, salinity, evaporation, and precipitation (Figure \ref{fig:clustering} $c$). Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations varied more strongly along NMDS 2. Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling but additionally depend on freshwater input from precipitation and surface runoff, showed greater variance along NMDS 1. Salinity, evaporation, and silicate concentration exhibited some explanatory power in the gradient forest analysis, although they were weaker predictors than SST, nitrate concentration, and climate indices (see Figure \ref{fig:GF}). This implies that the climate indices explain variability that more direct measurements of upwelling intensity do not capture.

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/FigureA3_GF_output_NOLAG_v1.pdf}
\caption{Gradient forest analysis of monthly community data at a genus level. (a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances. (b) Phytoplankton genera that could be predicted ranked according to the predictive power via the weighted importance. (c) Cumulative importance curves for the predictor variables. Black line shows cumulative importance across predictor range for the entire model and all genera. Dashed colored lines represent the responses of individual functional groups.}
\label{fig:sup:GFoutput_nolags}
\end{figure}

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/FigureA4_GF_output_Supplement_NOLAG_v1.pdf}
\caption{Cumulative importance curves for all variables not shown in the previous figure. Dashed black line represents the cumulative importance across the entire community, colored lines represent functional groups. Gradient forest runs where done on Genus-level cell counts, integrated over the top 100 meters.}
\label{fig:sup:GFoutput_nolags_extra}
\end{figure}

>>>>

Phytoplankton communities are integral to oceanic biogeochemical cycles and are sensitive indicators of climate-driven environmental variability.
Long-term time series capture this variability, allowing us to unravel the effects of environmental change on local communities. This study investigates changes in the phytoplankton community of the Cariaco Basin, off the coast of Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data. To investigate impacts of larger-scale climate variability, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both linked to variations in local weather and upwelling in the Cariaco Basin.
Cluster analysis identified two distinct community states separated by a shift from high to low upwelling conditions in 2004. In 2014, community and environmental conditions returned a state similar to that before the shift in 2004. No significant differences in diversity were observed between the two periods, however, the low-upwelling period was characterized by higher community evenness and lower species turnover.
To quantify effects of climate and bottom-up drivers on phytoplankton community changes, we employed gradient forest analysis and found the AMO index was the strongest predictor of changes in the community, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2. Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and suggests a possible cyclic (if not resilient) behavior of this tropical marine ecosystem under global change.
% Reviewer two criticizes mention of cyclic behaviour, since missing data might skew results. Need to check!
\end{abstract}

@@

% PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
The shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensity from 2004 to 2013 did not appear to represent a permanent regime change. Hydrographic data from 2014 to 2017, the final years of the CARIACO (CArbon Retention In A Colored Ocean) time series, indicated a return to strong upwelling conditions, while aggregated microscopy cell counts from 2014 and 2015 showed increased abundances of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}. Although the impact of large-scale climatic processes on the observed shifts in phytoplankton community composition remains unclear, we hypothesize that these changes were driven by decadal-scale changes in climatic factors. The Atlantic Multidecadal Oscillation (AMO) has emerged as an important factor influencing both the regional climate system and north Atlantic ecosystems \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean Sea, where a negative phase is generally correlated with warmer and drier conditions and a positive phase with colder and wetter conditions \cite{stephenson_changes_2014}. A negative AMO phase has also been linked to a southward shift in the ITCZ \cite{knight_climate_2006}. Such a southward shift of the ITCZ is correlated with increased wind-driven upwelling in the Cariaco Basin \cite{taylor_ecosystem_2012}. The El Niño Southern Oscillation (ENSO) is teleconnected to the Caribbean Sea, given that during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures increase and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. However, using the time series available only until 2010, \citeA{taylor_ecosystem_2012} found no significant correlation between MEI v.2 and in situ observations, and did not test for correlations with the AMO index.

% What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
\citeA{taylor_ecosystem_2012} noted that the observed trends of increasing SST, reduced upwelling intensity, and reduced chlorophyll \textit{a} between 1996 and 2010 could not be clearly identified as ``unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation''. To our knowledge, all analyses of phytoplankton community diversity in the Cariaco Basin do not include time series data beyond 2012, and no detailed investigations have yet been conducted on changes in diversity based on microscopy cell counts. Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 to 2017, 2) changes in the structure of the phytoplankton community and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community. Our analysis focuses exclusively on bottom-up drivers of community changes because these data are available throughout the entire time series and at a monthly resolution. Our study builds on and expands the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. In doing so, we aim to provide a more comprehensive assessment of long-term phytoplankton community dynamics in the Cariaco Basin, offering new insights into the local ecological responses to environmental variability across decadal scales.

@@

\subsection{Meteorological and climate data}
    %from Taylor et al. 2012
    In addition to the in situ data, we also examined meteorological and climatic data that are known to influence ocean conditions and marine ecosystems. Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}. We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the monthly averaged ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the CARIACO time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size). To signify the easterly wind, we used the negative of the u10 value in our analysis and, for simplicity, we refer to this variable as ``wind speed'' throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate of total precipitation, comprising all water that falls on Earth's surface as rain or snow. These variables were all available at monthly resolution across the entire time series. We used these variables, in addition to the climate indices, in the analysis of annual z-scores, in the NMDS clustering analysis to explain variance in the monthly community data, and as predictors of community change in the gradient forest analysis.

    In addition to the wind data, we examined two climatic indices, each of which provides information on the basin-scale climatic conditions and, consequently, the broad-scale forcing of marine ecosystems. The monthly Atlantic Multidecadal Oscillation (AMO) index (10-year low-pass) is calculated using area-averaged, detrended low-pass filtered North Atlantic SST anomalies (\ang{0}N-\ang{60}N) \cite{trenberth2023amo}. 
    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) index (MEI v.2) is defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five variables (sea level pressure, SLP; SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W) \cite{noaa_psl_meiv2}. In our analysis, the bi-monthly values were aligned with the tail month (i.e., December–January was assigned to January).

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using ``tidyverse'' packages \cite{wickham_welcome_2019}. All the codes we produced for data processing and data analysis are publicly available on GitHub \cite{benjamin_post_2025_16878343}.
    For all discrete depth measurements, the data were interpolated using the ``oce'' package \cite{kelley_oce_2023} with the adaptive ``unesco'' algorithm defined by the U.S. National Oceanographic Data Center \cite{johnson2006world}. Depth discrete data was generally available at depths of 1, 7, 15, 25, 35, 55, 75, and 100 m. The interpolated values were depth integrated to \qty{100}{\meter} and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis. Furthermore, the sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}. For phytoplankton microscopy data, individually identified taxonomic units (species or genus) were interpolated and integrated across depth, and finally grouped by genus via the sum of interpolated counts.

\subsection{Data analysis}    
    In the analysis, we set out to find if there was a structure in the distribution and temporal evolution of diversity metrics and community composition and, additionally, to identify potential bottom-up drivers of this change. 
    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores, $z = \frac{(x – \mu)}{\sigma}$, where $x$ is the yearly mean, $\mu$ is the mean and $\sigma$ is the standard deviation of all yearly means. For the analysis of annual z-scores, data from 1995, 2016, and 2017 were not included because insufficient coverage (i.e. more than six missing monthly samples) in these years prevented us from producing reliable yearly estimates. 
    %TALK MORE ABOUT MISSING DATA

@@

\subsection{AMO and upwelling-related variables drive community changes}
    % Figure 6
    To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a gradient forest analysis on the monthly microscopy cell counts interpolated over the top 100 meters and grouped by identified genus. Preliminary model runs were made to determine whether time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months. Figure \ref{fig:GF} $a$ shows the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity. The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentration scored the lowest in predicting changes within the community. 
    Gradient forest predicts shifts in the community by building a random forest model for each observed genus across the time series. The predictive power of the models for each genus is shown in Figure \ref{fig:GF} $b$. The microscopy data are biased toward larger cell sizes and, therefore, most of the predicted genus cell counts are diatoms. However, two haptophyte and three dinoflagellate genera rank relatively high. For the presented run a total of 89 genera (each observed more than 10 times over the time series) were used as indicator variables, of which 46 genera could be predicted using the given predictor variables, which corresponds to more than \qty{95}{\%} of counted cells (see Table \ref{tab:sup:GenusFuncGroups} for a complete list of identified genera). 
    Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the parameter range. The dashed black lines indicate cumulative importance across the entire community, and the solid colored lines indicate importance per phytoplankton group. 
    Cumulative importance curves from gradient forest analysis show how the importance of a predictor accumulates across its range of values. The slope of the curve at any point indicates where the strongest compositional turnover occurs—steep sections represent threshold values where small changes in the predictor correspond to large changes in community composition, while flat sections indicate regions where the predictor has little influence on community structure.
    For the AMO index, we see a clear pattern in which negative anomalies affect the community (slope is highest for negative values). This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes and cyanobacteria. Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range within which the community changed most, indicating these values as a threshold for community composition. Figure \ref{fig:sup:GFoutput_lags_extra} in Appendix shows response curves for all variables not shown here. Variables that show a pronounced threshold response are evaporation and salinity. For evaporation the threshold is at larger values of evaporation, which are most likely linked to periods of strong stratification, and salinity at both extreme ends of the range, signifying effects on cell counts of diatoms, dinoflagellates and cyanobacteria at strong upwelling and stratified conditions.

    \begin{figure}
    \noindent\includegraphics[width=\textwidth]{fig/Figure6_GF_output_FINAL_v2.pdf}
    \caption{Gradient forest analysis of monthly community data at a genus level. (a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances. (b) Phytoplankton genera that could be predicted ranked according to the predictive power via the weighted importance. (c) Cumulative importance curves for the predictor variables. The dashed black lines show cumulative importance across predictor range for the entire model and all genera. Solid colored lines represent the responses of individual functional groups.}
    \label{fig:GF}
    \end{figure}

@@

    % P3: Shift in Diversity does not match shift in biomass/cell counts, why?
    The temporal dynamics found in yearly mean genus richness, however, did not correspond directly to the observed clusters. Instead, there was a marked reduction from 1998 to 2009, after which richness recovered four years prior to 2014, when the community returned to levels characteristic of cluster 1. The same trend was evident, albeit to a lesser degree, in the other diversity metrics. Generally, mean annual anomalies in Shannon diversity and Pielou's evenness correlated with genus richness and exhibited a reduced yearly mean between 2000 and 2006, during which cell counts recorded an increase in nanoflagellates and a pronounced peak in cyanobacteria.
    We did not find a significant difference in diversity between the two clusters, either by Shannon index or by genus richness (Figures \ref{fig:zscore} and \ref{fig:divts}). However, we did observe a significant difference in Pielou's index between the two clusters. The results suggest that the phytoplankton community became more evenly distributed during the low upwelling conditions of cluster 2, which is consistent with the analysis of \citeA{pinckney_phytoplankton_2015} based on HPLC pigment data. This increase in evenness, quantified by the Pielou index, is most likely explained by the lack of strong upwelling events and the resulting reduced disturbance of the community. Under strong upwelling conditions, r-selected species are typically favored, leading to dominance by fewer taxa and consequently lower evenness. A decrease in compositional turnover was also observed in a decline in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$). As the productivity and biomass of phytoplankton communities are driven primarily by nutrients supplied through disturbance events, this is in line with the inverse relationship between evenness and community performance observed in previous studies \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.

@@

    The El Niño Southern Oscillation (ENSO) exhibits a strong teleconnection to the Caribbean Sea. During El Niño (indicated by a positive MEI v.2 index) sea surface temperatures rise, and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. This would imply a general reduction in upwelling strength during El Niño events in the Cariaco Basin, which has indeed been observed during the period from 1997 to 1998 \cite{romero_seasonal_2009}. The MEI v.2 is significantly positively correlated to local evaporation and salinity and negatively correlated to the AMO index, however it is not significantly correlated to local upwelling-related variables (see Figure \ref{fig:sup:correlation} in Appendix). The link here is less clear; while generally trade wind strength and SST are anti-correlated in the Caribbean Sea, positive trade wind anomalies have been observed during during El Niño conditions and the reverse during La Niña \cite{chang_coupled_2013}. \citeA{taylor_ecosystem_2012} found no strong correlation between MEI v.2 and other variables in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance using linear regression with a time lag of 12 months. In contrast, we found a relatively strong predictive power for changes in the phytoplankton community for the MEI v.2 index with a 4-month time lag. We also tested a 12-month time lag and found no significant improvement (Table \ref{tab:sup:GFlagtests} in Appendix).
    The threshold responses indicate a shift in the abundances of dinoflagellates for negative MEI v.2 anomalies, which represent La Niña conditions. Findings from Bayesian modeling showed that dinoflagellates responded positively to warmer, stratified conditions in the Cariaco Basin \cite{mutshinda_environmental_2013}. For cyanobacteria and haptophytes, we observed a pronounced threshold response to positive anomalies, demonstrating sensitivity to ENSO events. Diatoms do not exhibit a clear threshold response, but show a slightly stronger response in the positive range of the MEI v.2. Interestingly, positive anomalies in the MEI v.2, representing strong ENSO events, are mostly found in the high upwelling regime of cluster 1 (see Figure \ref{fig:clustering} $e$). This could be explained by previous findings, that strong El Niño events (as indicated by a positive MEI) increase the strength of currents in the Caribbean sea through modulating the easterly wind anomalies, thereby potentially increasing upwelling strength in the Cariaco Basin \cite{huang_enso_2023}.
    In particular, this intensification of the Caribbean Current may enhance the secondary upwelling that typically occurs between June and August \cite{rueda-roa_description_2018}, rather than affecting the primary wind-driven upwelling between December and May. Additionally, enhanced eddy activity near the Cariaco Basin during strong El Niño events can stimulate productivity and alter community composition through ventilation of sub-oxic waters below the sill \cite{astor_seasonal_2003}.

@@

    These results matched the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, as the scaling axis that separated the two clusters was NMDS 1, which co-varied most strongly with changes in AMO, MEI v.2, salinity, evaporation, and precipitation (Figure \ref{fig:clustering} $c$). Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations varied more strongly along NMDS 2. Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling but additionally depend on freshwater input from precipitation and surface runoff, showed greater variance along NMDS 1. The fact that salinity and the MEI v.2 align is consistent with the influence of ENSO on regional hydrology, as El Niño events typically result in reduced precipitation and drier conditions in the southern Caribbean, whereas La Niña years are associated with higher precipitation, leading to elevated silicate concentrations and lower surface salinity \cite{giannini_interdecadal_2001}. Salinity, evaporation, and silicate concentration exhibited some explanatory power in the gradient forest analysis, although they were weaker predictors than SST, nitrate concentration, and climate indices (see Figure \ref{fig:GF}). This implies that the climate indices explain variability that more direct measurements of upwelling intensity do not capture.

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/FigureA3_GF_output_NOLAG_v1.pdf}
\caption{Gradient forest analysis of monthly community data at a genus level for all predictors without any time lags. (a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances. (b) Phytoplankton genera that could be predicted ranked according to the predictive power via the weighted importance. (c) Cumulative importance curves for the predictor variables. The dashed black lines show cumulative importance across predictor range for the entire model and all genera. Solid colored lines represent the responses of individual functional groups.}
\label{fig:sup:GFoutput_nolags}
\end{figure}

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/FigureA4_GF_output_Supplement_NOLAG_v1.pdf}
\caption{Cumulative importance curves for all variables not shown in the previous figure, showing the gradient forest analysis performed with all predictors without any time lags. Dashed black line represents the cumulative importance across the entire community. Solid colored lines represent functional groups. Gradient forest runs where done on Genus-level cell counts, integrated over the top 100 meters.}
\label{fig:sup:GFoutput_nolags_extra}
\end{figure}
